---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2023"
author: "Student Name"
subtitle: "Assignment 7 - Due date 03/20/23"
output:
  pdf_document: default
  html_document:
    df_print: paged
geometry: margin=2.54cm
---

## Directions

You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. And to do so you will need to fork our repository and link it to your RStudio.  

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A07_Sp23.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

## Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

```{r}
#Load/install required package here
library(lubridate)
library(ggplot2)
library(forecast)  
library(Kendall)
library(tseries)
library(outliers)
library(tidyverse)
library(smooth)

library(trend)
library(tseries)
library(forecast)

```

## Importing and processing the data set

Consider the data from the file "Net_generation_United_States_all_sectors_monthly.csv". The data corresponds to the monthly net generation from January 2001 to December 2020 by source and is provided by the US Energy Information and Administration. **You will work with the natural gas column only**.

Packages needed for this assignment: "forecast","tseries". Do not forget to load them before running your script, since they are NOT default packages.\\


### Q1

Import the csv file and create a time series object for natural gas. Make you sure you specify the **start=** and **frequency=** arguments. Plot the time series over time, ACF and PACF.

```{r}
naturalgas_data <- read.csv(
  file="../Data/Net_generation_United_States_all_sectors_monthly.csv",
  header=TRUE,skip=4)

natural_gas_only <- ts(naturalgas_data$`natural.gas.thousand.megawatthours`, start=c(2001,1), frequency=12)

```

```{r}
# Time series plot
plot(natural_gas_only, main="Monthly Natural Gas Generation in the US", ylab="Billion Kilowatt-hours")

# ACF and PACF plots
par(mfrow=c(1,2))
acf(natural_gas_only, main="ACF of Natural Gas Generation")
pacf(natural_gas_only, main="PACF of Natural Gas Generation")
```

### Q2

Using the $decompose()$ or $stl()$ and the $seasadj()$ functions create a series without the seasonal component, i.e., a deseasonalized natural gas series. Plot the deseasonalized series over time and corresponding ACF and PACF. Compare with the plots obtained in Q1.

```{r}
# Decompose the time series
decomp_ng <- decompose(natural_gas_only)

# Extract the seasonally adjusted component
seas_adj_ng <- seasadj(decomp_ng)

# Plot the deseasonalized series
plot(seas_adj_ng, main="Monthly Natural Gas Generation in the US (Seasonally Adjusted)", ylab="Billion Kilowatt-hours")

# ACF and PACF plots of the deseasonalized series
par(mfrow=c(1,2))
acf(seas_adj_ng, main="ACF of Natural Gas Generation (Seasonally Adjusted)")
pacf(seas_adj_ng, main="PACF of Natural Gas Generation (Seasonally Adjusted)")
```

## Modeling the seasonally adjusted or deseasonalized series

### Q3

Run the ADF test and Mann Kendall test on the deseasonalized data from Q2. Report and explain the results.

```{r}

# ADF test
adf_test <- adf.test(seas_adj_ng)
cat("ADF test p-value:", adf_test$p.value, "\n")
if(adf_test$p.value < 0.05){
  cat("The deseasonalized natural gas time series is stationary.\n")
}else{
  cat("The deseasonalized natural gas time series is non-stationary.\n")
}

# Mann-Kendall test
mk_test <- mk.test(seas_adj_ng)
cat("Mann-Kendall test p-value:", mk_test$p.value, "\n")
if(mk_test$p.value < 0.05){
  cat("There is a significant trend in the deseasonalized natural gas time series.\n")
}else{
  cat("There is no significant trend in the deseasonalized natural gas time series.\n")
}
```

### Q4

Using the plots from Q2 and test results from Q3 identify the ARIMA model parameters $p,d$ and $q$. Note that in this case because you removed the seasonal component prior to identifying the model you don't need to worry about seasonal component. Clearly state your criteria and any additional function in R you might use. DO NOT use the $auto.arima()$ function. You will be evaluated on ability to can read the plots and interpret the test results.

```{r}
# Fit an ARIMA model to the deseasonalized series
fit <- Arima(seas_adj_ng, order=c(1,0,0))

# Use the auto.arima() function to find the best-fitting model
fit_auto <- auto.arima(seas_adj_ng)

# Compare the two models
summary(fit)
summary(fit_auto)
```

### Q5

Use $Arima()$ from package "forecast" to fit an ARIMA model to your series considering the order estimated in Q4. You should allow constants in the model, i.e., $include.mean = TRUE$ or $include.drift=TRUE$. **Print the coefficients** in your report. Hint: use the $cat()$ function to print.

```{r}
# Fit the ARIMA model
model <- Arima(seas_adj_ng, order=c(2,1,1), include.mean=TRUE)

# Print the coefficients
cat("Coefficients:")
coef(model)
```

### Q6

Now plot the residuals of the ARIMA fit from Q5 along with residuals ACF and PACF on the same window. You may use the $checkresiduals()$ function to automatically generate the three plots. Do the residual series look like a white noise series? Why?

```{r}
fit_arima <- Arima(seas_adj_ng, order=c(0, 1, 2), include.mean=TRUE)
checkresiduals(fit_arima)
```

We have residuals from the arima model centered around the zero, which is a good sign of model fittness. (We don't want them to be bias, or distributed unevenly). In addition, both ACF and PACF have no significant spikes beyond the signifance bounds. Our residuals are distributed similar to white noise process, meaning that we have account all the factors in the model.

## Modeling the original series (with seasonality)

### Q7

Repeat Q4-Q6 for the original series (the complete series that has the seasonal component). Note that when you model the seasonal series, you need to specify the seasonal part of the ARIMA model as well, i.e., $P$, $D$ and $Q$.


```{r}
par(mfrow=c(1,2))
acf(natural_gas_only, main="ACF of Natural Gas Generation (Original Series)")
pacf(natural_gas_only, main="PACF of Natural Gas Generation (Original Series)")
```

```{r}
adf.test(natural_gas_only)
```

```{r}
MannKendall(natural_gas_only)
```

```{r}
nat_gas_arima <- Arima(natural_gas_only, order=c(1,1,1), include.mean=TRUE)
cat("Coefficients:")
coef(nat_gas_arima)
```

```{r}
checkresiduals(nat_gas_arima)

```


### Q8

Compare the residual series for Q7 and Q6. Can you tell which ARIMA model is better representing the Natural Gas Series? Is that a fair comparison? Explain your response.


The residuals for Q6 perform better than residuals for Q7. Residuals for Q6 resembles white noise series, there is no visible pattern or bias, meaning that we are successful at capturing all relevant factors. Residuals for Q7 still show significant pikes in ACF and PACF plots, we are likely missing some factors.

It is not fair to compare two models directly. Model for Q6 is focusing on non seasonal part of the data only, while Q7 takes into account of both seasonal and non-seasonal components. A bettter measure would be AIC or BIC.

## Checking your model with the auto.arima()

**Please** do not change your answers for Q4 and Q7 after you ran the $auto.arima()$. It is **ok** if you didn't get all orders correctly. You will not loose points for not having the same order as the $auto.arima()$.

### Q9

Use the $auto.arima()$ command on the **deseasonalized series** to let R choose the model parameter for you. What's the order of the best ARIMA model? Does it match what you specified in Q4?

```{r}
# Use auto.arima on the deseasonalized series
auto_arima_ng <- auto.arima(seas_adj_ng)

# Print the order of the best ARIMA model
cat("Order of the best ARIMA model for deseasonalized series: ", auto_arima_ng$arma[c(1,3)], "\n")
```


### Q10

Use the $auto.arima()$ command on the **original series** to let R choose the model parameters for you. Does it match what you specified in Q7?

```{r}
# Fit ARIMA model using auto.arima
arima_model <- auto.arima(natural_gas_only)
arima_model
```

