---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2023"
subtitle: "Assignment 2 - Due date 02/03/23"
author: "Caroline Ren"
output: pdf_document
geometry: margin=2.54cm
---

## Submission Instructions

You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github.   

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A02_Sp23.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

## R packages

R packages needed for this assignment:"forecast","tseries", and "dplyr". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
#Load/install required package here
library(forecast)
library(tseries)
library(dplyr)
```


## Data set information  

Consider the data provided in the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx" on our **Data** folder. The data comes from the US Energy Information and Administration and corresponds to the December 2022 Monthly Energy Review. The spreadsheet is ready to be used. You will also find a $.csv$ version of the data "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source-Edit.csv". You may use the function $read.table()$ to import the $.csv$ data in R. Or refer to the file "M2_ImportingData_CSV_XLSX.Rmd" in our Lessons folder for functions that are better suited for importing the $.xlsx$.

```{r}
#Importing data set
energy_data <- read.csv(file="../Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source-Edit.csv", header = TRUE, dec = ".", sep=",",stringsAsFactors = TRUE) 

#check that colnames are already stored because we deleted the units line so we were able to automatically import it
colnames(energy_data)
head(energy_data)
```


## Question 1
You will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. Use the command head() to verify your data.

```{r}
#Use dplyr select
energy_data1 <- energy_data %>% select(1,4,5,6) 

nenergy <- ncol(energy_data1)-1
nobs <- nrow(energy_data1) 

head(energy_data1)

```


## Question 2
Transform your data frame in a time series object and specify the starting point and frequency of the time series using the function ts().

```{r}
ts_energy_data <- ts(energy_data1[,2:4],frequency=12,start=c(1973,1))

ts_energy_data
```


## Question 3
Compute mean and standard deviation for these three series.

```{r}
mean_biomass <- mean(energy_data1$Total.Biomass.Energy.Production, na.rm=TRUE)
mean_renewbale <- mean(energy_data1$Total.Renewable.Energy.Production, na.rm=TRUE)
mean_hydro <- mean(energy_data1$Hydroelectric.Power.Consumption, na.rm=TRUE)

sd_biomass <- sd(energy_data1$Total.Biomass.Energy.Production, na.rm=TRUE)
sd_renewable <- sd(energy_data1$Total.Renewable.Energy.Production, na.rm=TRUE)
sd_hydro <- sd(energy_data1$Hydroelectric.Power.Consumption, na.rm=TRUE)

mean_biomass
mean_renewbale
mean_hydro
sd_biomass
sd_renewable
sd_hydro

```


## Question 4
Display and interpret the time series plot for each of these variables. Try to make your plot as informative as possible by writing titles, labels, etc. For each plot add a horizontal line at the mean of each series in a different color.

```{r}
plot(energy_data1[,"Total.Biomass.Energy.Production"],type="l",col="blue",xlab="Month over years",ylab="Production",main="Total Biomass Energy Production") 

#Additional - Suppose you want to add a line with the mean
abline(h=mean(energy_data1[,"Total.Biomass.Energy.Production"]),col="red")

```

```{r}
plot(energy_data1[,"Total.Renewable.Energy.Production"],type="l",col="blue",xlab="Month over years",ylab="Production",main="Total Renewable Energy Production") 

#Additional - Suppose you want to add a line with the mean
abline(h=mean(energy_data1[,"Total.Renewable.Energy.Production"]),col="red")

```


```{r}
plot(energy_data1[,"Hydroelectric.Power.Consumption"],type="l",col="blue",xlab="Month over years",ylab="Production",main="Hydroelectric Power Consumption") 

#Additional - Suppose you want to add a line with the mean
abline(h=mean(energy_data1[,"Hydroelectric.Power.Consumption"]),col="red")

```

The annual trend is more obvious than seasonal trend for biomass and renewable energy production. Both two show increasing trend over years. It is more likely that biomass show a polynomial increase in annual trend, and renewable shows a linear increase in annual trend. For biomass, the seasonal pattern was more obvious in some of the year over other, likely due to some external events that happened. For renewable energy, the seasonal fluctuations are relatively constant throughout the year. For hydroelectric power, seasonal trend is more obvious than annual trend.

## Question 5
Compute the correlation between these three series. Are they significantly correlated? Explain your answer.

```{r}
cor(ts_energy_data) 
```
By looking at the correlation coefficients, the coefficients are highest between biomass and renewable (0.92). The correlation between biomass and hydroelectric or renewable and hydroelectric are very weak, which has a correlation coefficient of (-0.30 and -0.10) respectively.

## Question 6
Compute the autocorrelation function from lag 1 up to lag 40 for these three variables. What can you say about these plots? Do the three of them have the same behavior?

```{r acf}
Biomass_acf=acf(ts_energy_data[,1],lag.max=40, type="correlation", plot=TRUE)
Renewable_acf=acf(ts_energy_data[,2],lag.max=40, type="correlation", plot=TRUE)
Hydro_acf=acf(ts_energy_data[,3],lag.max=40, type="correlation", plot=TRUE)


```

Both biomass and renewable’s acfs have greater correlation with more recent lags, which is lost over time, which is expected since distant lags are likely contribute less to the y.

For hydroelectric, the correlation decreases over time, which is expected. However the value spikes and then decreases again in cycles. These are likely due to seasonal effects.


## Question 7
Compute the partial autocorrelation function from lag 1 to lag 40 for these three variables. How these plots differ from the ones in Q6?

```{r}
Biomass_pacf=pacf(ts_energy_data[,1],lag.max=40, plot=TRUE)
Renewable_pacf=pacf(ts_energy_data[,2],lag.max=40, plot=TRUE)
Hydro_pacf=pacf(ts_energy_data[,3],lag.max=40, plot=TRUE)
```

PCF removes the intermediate values and check the correlation between lag and y directly. Both biomass and renewable’s partial acf shows no significant correlations between lags and observations for the most part, despite some outliers.

PCF for hydroelectric has a really high peak in the earlier lags, even after removing intermediate values. Seasonal effect plays a strong role here, seasons likely affect values in the previous season, for example, the current hydroelectric production in spring is correlated with the hydroelectric production in last spring.


